---
title: "motif interactions"
output: html_document
date: "2026-04-05"
---

```{r}
suppressPackageStartupMessages({
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(betterChromVAR)
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(TFBSTools)
  library(sechm)
  library(ggplot2)
  library(BiocParallel)
  library(epiwraps)
})
```

We get some motifs:

```{r}
motifs <- query(MotifDb, c("HOCOMOCOv13"))
motifs <- motifs[grepl("A$|B$",names(motifs))]
motifs <- do.call(PWMatrixList, convert_motifs(motifs, "TFBSTools-PWMatrix"))
names(motifs) <- sapply(motifs, \(x) x@name)
motifs <- motifs[!duplicated(names(motifs))]
```


# We load the NR3C1 activation dataset

We load the peak counts

```{r}
se <- readRDS("NR3C1.rds")
se <- se[rowSums(assay(se))>0,]
se$group <- gsub("_?[0-9]\\.bam","",colnames(se))
```

Find motif matches:

```{r, eval=FALSE}
moi <- motifmatchr::matchMotifs(motifs, e, genome=BSgenome.Hsapiens.UCSC.hg38)
saveRDS(moi, file="NR3C1_moi.rds")
```

```{r, include=FALSE}
moi <- readRDS("NR3C1_moi.rds")
```

# Screening for modulators of the GCR response

```{r}
# get deviations for motif intersections:
idev <- getInteractionsDeviations(se, moi, bait="GCR", minCount = 10)
# test for significant interactions:
res <- discoverMotifInteractions(idev, "group")
head(res)
```

The reported p-values are relative to the difference (column 'difference') in response to the activation (i.e. difference between the two groups) in sites harboring both motifs (GCR + the listed motif), versus sites harboring GCR more generally. In other words, a positive 'difference' means that
sites having both motifs respond more than average GCR sites, while a negative difference means that
they respond less. (The magnitude of this difference can be directly compared to the magnitude of the difference in GCR deviation.)
The percentage of overlap between the GCR and the candidate motif (as a fraction of GCR sites) is also reported, as well as the Jaccard index. This is important because a very high overlap suggests that the two motifs are probably very similar to each other (as for the example of MCR, see below), while a very small overlap indicates that the interaction concerns a small fraction of the sites.

We can visualize the results in this way:


```{r, fig.width=12, fig.height=4}
p1 <- ggplot(res, aes(difference, -log10(adj.P.Val), label=motif, size=overlap.N, colour=overlap.jaccard)) + 
  geom_point(alpha=0.2) + 
  ggrepel::geom_text_repel(data=head(res,10),
                           min.segment.length=0, max.overlaps=30) + 
  scale_size_continuous(breaks=c(min(res$overlap.N),500,1500)) + 
  scale_color_viridis_c(option="A", trans="sqrt") + theme_bw() + 
  labs(x="Difference in response", size="Number\nof sites", 
       colour="Jaccard\nwith bait", title="Interactions with GCR")
selmo <- head(row.names(res),10)
p2 <- cowplot::plot_grid(plotlist = lapply(motifs[selmo], \(x){
  view_motifs(x, show.positions=FALSE) + theme_void() + theme(legend.position = "none")
  }), ncol=1, labels=selmo, scale=0.95, hjust = 0, label_size = 11, label_fontface = 0)
(pres <- cowplot::plot_grid(p1, p2, nrow=1, scale=c(0.95, 0.9), labels="AUTO"))
```


The top candidates are the well-known interactors of GCR, the FOS/JUN family members forming the AP-1 complex. We also find RELB, which forms a complex with GCR. Beside these, we  mostly have 3 non-redundant candidates: ZN133, PO4F1, and CTCFL.
We can now look at them in more detail.

```{r, fig.width=12, fig.height=8}
# we pre-compute background data for speed:
bg <- getBackgroundBins(se)
bg <- computeBackgrounds(se, bg)
# for each potential partner (here looking at the top 3 non-redundant), we look at the 
# interaction in more detail, and melt the values into a data.frame for plotting
candidates <- c("JUNB", "ZN133", "PO4F1", "RELB", "CTCFL")
dd <- bplapply(candidates, BPPARAM=MulticoreParam(min(5,length(candidates)), progress=TRUE), \(mo){
  dev <- exploreMotifInteraction(se, motifs[c("GCR", mo)], bg = bg,
                                 genome=BSgenome.Hsapiens.UCSC.hg38)
  d <- meltSE(dev, row.names(dev))
  d$feature <- factor(d$feature, row.names(dev))
  d$query <- mo
  d
})
dd <- dplyr::bind_rows(dd)
dd$query <- factor(dd$query, unique(dd$query))
p3 <- ggplot(dd, aes(feature, norm)) + geom_hline(yintercept=0, linetype="dashed") + 
  geom_boxplot(aes(fill=group)) + facet_wrap(~query, scale="free", nrow=2) + 
  theme_bw() + labs(x="Site type", y="Adjusted z-scores") + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
        legend.position = c(0.8, 0.15), legend.justification = c(0.5, 0.5)
  )
p3
```

The JUNB candidate is a perfect example of synergy: JUNB sites by themselves do not change in accessibility, while the GCR+JUNB- sites change slightly less than the average GCR+ sites, and the GCR+JUNB+ sites change much more, in a way that decays with distance between the two motifs. RELB shows a similar pattern, although decaying faster with distance, consistent with a direct interaction with GCR.


```{r, fig.width=10, fig.height=12}
pdf("Figure_motifInteraction.pdf", width=10, height=12)
cowplot::plot_grid(pres, p3, nrow=2, scale=c(1,0.95),
                   labels=c(NA, "C"), rel_heights=c(1.1,2))
dev.off()
```


```{r}
sessionInfo()
```

